---
title: "Posture effects (Seated vs Semi-standing vs Standing)"
output: html_document
---

# 1. Libraries and Data Import

This section installs missing packages if needed, then loads the normalized Vicon dataset.


```{r setup, include=FALSE}
required_packages <- c(
  "dplyr",
  "tidyr",
  "readxl",
  "ggplot2",
  "lme4",
  "lmerTest",
  "emmeans",
  "here",
  "stringr",
  "afex"
)

installed <- rownames(installed.packages())
to_install <- setdiff(required_packages, installed)

if (length(to_install) > 0) {
  message("Installing missing packages: ", paste(to_install, collapse = ", "))
  install.packages(to_install, repos = "https://cloud.r-project.org")
}

invisible(lapply(required_packages, library, character.only = TRUE))

```

```{r}

# Choose dataset version ("test" or "raw")
MODE <- "test"  # Change to "raw" for full dataset

# Define imput path
in_path <- here("results", MODE, "vicon_MQI_wrists_head_normByShoulders.xlsx")

# Ensure file exists
stopifnot(file.exists(in_path))
# Load data
df <- read_excel(in_path)
# Inspect variable names
names(df)
```
# 2. Extract Posture and Dyad Information

We extract posture condition and dyad ID from the video identifier.

```{r}
df <- df %>%
  mutate(
    # Extract posture condition from video_id
    posture = case_when(
      str_detect(video_id, "^SEATED")   ~ "SEATED",
      str_detect(video_id, "^SEMI")     ~ "SEMI",
      str_detect(video_id, "^STANDING") ~ "STANDING"
    ),
     # Extract dyad ID (e.g., D01, D02...)
    dyad = str_extract(video_id, "D\\d+")
  )

table(df$posture)

```

# 3. Reshape Data to Long Format

We reshape wrist and head variables separately.
Each participant (P1 and P2) is treated as an independent subject.

```{r}
# WRIST movement (normalized)
df_wrist_long <- df %>%
  select(dyad, posture,
         P1_MQI_WRISTS_norm,
         P2_MQI_WRISTS_norm) %>%
  pivot_longer(
    cols = c(P1_MQI_WRISTS_norm, P2_MQI_WRISTS_norm),
    names_to = "participant",
    values_to = "value"
  ) %>%
  mutate(
    subject = paste0(dyad, "_", participant),
    value = as.numeric(value)
  )
# HEAD movement (normalized)
df_head_long <- df %>%
  select(dyad, posture,
         P1_MQI_HEAD_norm,
         P2_MQI_HEAD_norm) %>%
  pivot_longer(
    cols = c(P1_MQI_HEAD_norm, P2_MQI_HEAD_norm),
    names_to = "participant",
    values_to = "value"
  ) %>%
  mutate(
    subject = paste0(dyad, "_", participant),
    value = as.numeric(value)
  )

```

# 4. Repeated Measures ANOVA (Greenhouse–Geisser correction)

We test the effect of posture on normalized movement.
Sphericity violations are corrected using Greenhouse–Geisser adjustment.

```{r}
library(afex)
# WRIST ANOVA
anova_wrist <- aov_ez(
  id = "subject",
  dv = "value",
  within = "posture",
  data = df_wrist_long,
  anova_table = list(correction = "GG", es = "pes")
)

anova_wrist
# HEAD ANOVA
anova_head <- aov_ez(
  id = "subject",
  dv = "value",
  within = "posture",
  data = df_head_long,
  anova_table = list(correction = "GG", es = "pes")
)

anova_head
```
# 5. Post-hoc Comparisons (Bonferroni correction)

Pairwise comparisons between posture conditions.

```{r}
emmeans(anova_wrist, pairwise ~ posture, adjust = "bonferroni")
emmeans(anova_head, pairwise ~ posture, adjust = "bonferroni")
```

